---
title: "Predicting heart disease with classification trees"
author: "Marko Orescanin"
date: "Nov 11, 2018"
output: html_document
---

<!-- change echo=FALSE to echo=TRUE to show code -->
```{r global_options, include=FALSE}
knitr::opts_chunk$set(prompt=TRUE, comment="", echo=TRUE)
```



The following code will load the libraries that will be used in order to create a classification tree for heart disease.

```{r collapse=TRUE, warning=FALSE}
library(rpart)
library(rpart.plot)
library(maptree)
# the following utility files can be found attached to the assignment
source("https://raw.githubusercontent.com/grbruns/cst383/master/lin-regr-util.R")
source("https://raw.githubusercontent.com/grbruns/cst383/master/class-util.R")
```

### Reading and preprocessing the data



First, we will preprocess the data.

```{r}
heart = read.table("https://raw.githubusercontent.com/grbruns/cst383/master/heart.dat", quote = "/")
names(heart) <- c("AGE", "SEX", "CHESTPAIN", "RESTBP", "CHOL",
                  "SUGAR", "ECG", "MAXHR", "ANGINA", "DEP", "EXERCISE", "FLUOR",
                  "THAL", "OUTPUT")
names(heart) = tolower(names(heart))

# convert output to factor
heart$output = factor(heart$output)
```

### Data exploration



Here are a couple of plots showing the relationship of cholestrol, restbp, maxhr, and age with heart disease.

```{r test}
par(mfrow= c(1,2))
plot(heart$chol, heart$output, main = "Heart Disease over cholesterol level", ylab = "Heart Disease", xlab = "Cholesterol", col = "red")
plot(heart$restbp, heart$output, main = "Heart Disease over RestBP", ylab = "heart disease", xlab = "restbp", col = "red")
par(mfrow= c(1,2))
plot(heart$maxhr, heart$output, main = "Heart Disease over MaxHR", ylab = "heart disease", xlab = "maxhr", col = "red")
plot(heart$age, heart$output, main = "Heart Disease over Age", ylab = "heart disease", xlab = "age", col = "red")
```

### Building a classification tree



The following code creates test and training data. The training data is used to create the classification tree. Inlcuded is a summary of the fit of the tree.

```{r}
# training and test sets
set.seed(132)
split = split_data(heart)
tr_dat = split[[1]]
te_dat = split[[2]]
heart$output = factor(ifelse(heart$output == 1, "Negative", "Positive"))
fit = rpart(output ~ maxhr + age, data = tr_dat, method = "class")
summary(fit)
prp(fit, extra=106, varlen=-10, main="classification tree for heart disease",
box.col=c("palegreen", "pink")[fit$frame$yval])
```

### Classifying test data



```{r}
predicted = predict(fit, te_dat, type="class")
predicted2 = predict(fit, tr_dat, type="class")
```

### Assessing the model



Here is a confusion matrix to show the accuracy of the tree, and the average of how many of the predicted values were actually heart disease.

```{r}
actual = te_dat$output
table(actual, predicted)
mean(actual == predicted)
```














